\chapter*{Acknowledgements}
\thispagestyle{empty}

I am indebted to Vincent Detours for the opportunity to realize this
dissertation in his research group, for his patience and his stern commitment to
individual emancipation.  Jacques Dumont was throughout this thesis a vibrant
source of motivation and a pillar for intellectual exchange.  I am also grateful
to Carine Maenhaut and Pierre Roger for supervising the progression of this work
at large.

% Long gone are the days when scientific research was the endeavour of lonely men
% and women.
This work could not have been without the wisdom and support of the many
colleagues I had the pleasure to meet during these formative years.  I am
particularly beholden to the the warm companionship of Maxime Tarabichi, Danai
Fimereli and David Gacquer.  Soazig le Pennec, Tomasz Konopka, David Weiss
Solís, Wilma van Staveren, Genevieve Dom, Soetkin Verstheye, Sébastien Floor,
Aline Hebrant, Aline Antoniou, Jaime Pita, Raquel Ramos, Robert Opitz, Luca
Tiberi, Benjamin Beck, Roxane van Heurck, Gabriele Zoppoli, David Brown, Germain
Mazères, Fredérique Savagner and Christophe Trésallet are among the many I had
the privilege to work and learn with during this doctorate.  Many thanks as well
to Danièle Leemans-De Vos, Genevieve Dalle and Joelle Sente for their invaluable
assistance.  I hope you won't mind if I call you my friends.

The years I spent working on this thesis will always be tied with my memories of
Brussels.  My dear friends Kenneth, Irina, Andrzej, Ivan, Florent, Ricardo,
Seyne, Kate, Yanna, Iris, Mathilde, Chiara, David, Eric, Nereida, Elvis, Steven,
Fillipo, and all the lost souls at the Lord Byron and at the Cobra are also part
of this.

Portugal is another home of mine.  In spite of the distance, the fingerprints of
my amigos Luísa, Pedro, Xana, Catarina, Bruno, Cristiana, Jorge e Armando can
be found all throughout this document.

I must not forget to mention Ana Sofia Rocha, without whom this thesis would not
have begun nor ended, and Karen Haast, who held my hand throughout so much of
it.

Finally, my deepest appreciation goes to the unwavering commitment and support
of my family.  To Celina, Gabriel, and very specially to my mom and dad---muito
obrigado por tudo.

\section*{Financial Support}
\label{sec:financial-support}
This work received the financial support of \emph{l'agence Wallonie-Bruxelles
  International}, of the \emph{Hoguet Foundation}, and of the \emph{Fonds David
  \& Alice van Buuren}.

\clearpage